---
title: "Calculate and plot boundaries"
author: "Johanna Braun"
date: "`r Sys.Date()`"
output:
  html_document: default
  # pdf_document: default
  md_document:
    variant: gfm
knit: (function(inputFile, encoding){
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
vignette: >
  %\VignetteIndexEntry{Calculate and plot boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Calculate and plot boundaries** `r if (!knitr::is_latex_output()) {"&#127939;"}`
describes the calculation and plotting of planetary boundary statuses at
different spatial scales based on LPJmL. It aims to simplify the post-processing
of LPJmL output data and to make the calculations and visualizations transparent
and flexible.

\newline

## Overview

The calculations rely on the output of LPJmL simulations. 
To configure and submit LPJmL simulations, the LPJmL Runner from the lpjmlkit
package can be used (https://github.com/PIK-LPJmL/lpjmlkit). In doing so, it is
important to ensure that all necessary output variables are written out (at the
needed spatial resolution). To do so, `boundaries` provides a function to list
all needed LPJmL input and output variables (see `?list_outputs`). These have to
be included in the LPJmL configuration file (`lpjml_config.cjson`).
Example output files can be downloaded from
https://doi.org/10.5281/zenodo.12171096.

For the post-processing, two working steps are required:
Calculate the boundaries (1) and plot them (2).
In a seperate step, a global validation table can be created to compare the
simulated key variables with global literature estimates (3). 

#### **1. `r if (!knitr::is_latex_output()) {"&#128203;"}` Calculate boundaries**\
  \
  Define in `calc_status` which boundaries should be calculated, at which
  spatial scale and for which timeframe, as well as additional parameters such
  as the calculation approach to use or the boundary and high risk values
  (see `?calc_status` *for more information*). This is the core wrapper function
  that calls the functions to calculate the individual boundaries.
  To change parameters that are specific to individual boundaries (e.g. the
  minimum tree cover to define forest biomes for the land system change
  boundary, `tree_cover_thresholds`), please refer to the documentation of the
  individual boundary status functions (see `?[boundary_name]_status`). 
  
  The function returns a list with the status of the contol variable(s) with
  attributes, such as the boundary and high risk values, the unit, and
  the spatial resolution.

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  global_status <- calc_status(
    boundary = c("bluewater", "greenwater", "lsc", "nitrogen", "biosphere"),
    config_scenario = config_path_scenario,
    config_reference = config_path_reference,
    spatial_scale = "global",
    path_baseline = output_path_baseline,
    time_span_scenario = as.character(1901:2017),
    time_span_reference = as.character(1500:1699),
    time_span_baseline = as.character(1901:2017),
    time_series_avg = 1, # computation of timeseries, no moving average
    approach = list("bluewater" = "porkka2024",
                    "greenwater" = "porkka2024",
                    "nitrogen" = "schulte_uebbing2022")
  )  
  ```

\newpage

#### **2. `r if (!knitr::is_latex_output()) {"&#9997;"}` Plot boundary status(es)**\
  \
  The status of the control variable(s) can now be visualized using the
  `plot_status` function, by transferring the output of the `calc_status`
  function. Depending on the spatial scale, the function applies different
  plotting functions (maps for `grid` and `regional` scale; timeseries plots for
  `global` scale). See `?plot_status` *for more information*, as well as the
  individual plotting functions called by `plot_status` for the different
  spatial scales (`status_map` for `grid` and `regional` scale;
  `status_stylized` (radial plot) or `status_global` (Cartesian coordinate
  system) for `global` scale).
  For some plots, the legend is not automatically plotted, but can be plotted
  separately using the `status_legend` function.

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  plot_status(global_status, filename = "./global_status.png")
  status_legend(filename = "./legend.png")
  ```

#### **3. `r if (!knitr::is_latex_output()) {"&#128269;"}` Create global validation table**\
  \
  The function `validate_simulation` compares relevant key variables that impact
  the control variable status, such as (i) inputs to the LPJmL model (e.g.
  nitrogen fertilization) and (ii) processed LPJmL outputs (e.g. irrigation
  water consumption), as well as (iii) the control variable status values
  themselves with global literature estimates. The parameter settings should
  match those chosen for application of the `calc_status` function at the global
  level, except for `time_span_scenario`, which should match the timespans/years
  of the literature estimates (see `./inst/extdata/global_validation_data.csv`
  for an overview of included literature data). The computed results will
  be averaged over this timespan. 
  See `?validate_simulation` *for more information*.

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  validation_table <- validate_simulation(
    config_scenario = config_path_scenario,
    config_reference = config_path_reference,
    path_baseline = output_path_baseline,
    time_span_scenario = as.character(2005:2014),
    time_span_reference = as.character(1500:1699),
    filename = "./validation_table.csv",
    approach = list("bluewater" = "porkka2024",
                    "greenwater" = "porkka2024",
                    "nitrogen" = "schulte_uebbing2022")
  )
  ```


#### **miscellaneous** \
  \
  More helpful functions that come with the package are:

  * `classify_biomes` to determine spatially-explicit biome extents based on the
    foliar projected cover of different plant functional types simulated in 
    LPJmL, the annual mean temperature and elevation (inputs to LPJmL). This
    function is used, amongst others, to calculate the land system change
    boundary status but can also be used independently.

  * `plot_biomes` to visualize the spatially-explicit biome extents calculated
    with `classify_biomes`.

  * `calc_efrs` to calculate environmental flow requirements based on different
    defined approaches based on LPJmL discharge output data. This is used for 
    the blue water boundary status at the gridded level but can be relevant also
    for other applications.
  
  * `as_risk_level` to convert the status of a control variable to a risk level
    based on the boundary and high risk values. If the risk level is not to be
    plotted but needed for further analysis, this function can be directly
    applied to the output of `calc_status`.
  
  For an overview on all implemented approaches, the respective control
  variables, units, the default boundary & high risk values, as well as all
  needed LPJmL outputs, see `./extdata/metric_files.yml`.


&nbsp;
&nbsp;

\newpage

## Usage

```{r setup, echo = TRUE, eval = FALSE, highlight = TRUE}
library(boundaries)
# Why tibble? -> https://r4ds.had.co.nz/tibbles.html
# Tibbles also provide a better overview of the data and directly show the type
# of each column, which is very important for integer/floating point values.
# TODO any other packages needed? biospheremetrics, tidyr, lpjmlkit?)

library(tibble)

# set paths to lpjml configuration files and outputs
path_configurations <- "./lpjml/configurations/"
config_path_scenario <- paste0(path_configurations, "config_scenario.json")
config_path_reference <- paste0(path_configurations, "config_reference.json")
path_outputs <- "./lpjml/output/"

plotpath <- paste0("./R/plots/")

  validation_table <- validate_simulation(
    config_scenario = config_path_scenario,
    config_reference = config_path_reference,
    path_baseline = output_path_baseline,
    time_span_scenario = as.character(2005:2014),
    time_span_reference = as.character(1500:1699),
    filename = "./validation_table.csv",
    approach = list("bluewater" = "porkka2024",
                    "greenwater" = "porkka2024",
                    "nitrogen" = "schulte_uebbing2022")
  )

```

&nbsp;
&nbsp;


### Global status timeseries\

#### **Example** *Global status calculation with moving average*  

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
### calc status at the global level

# define analysis and reference period
time_span_scenario <- as.character(1901:2017)
time_span_reference <- as.character(1500:1699)

# define the number of years to average over, for a moving average; if set to
# 1, no moving average is calculated
nyear_window <- 5

global_status <- calc_status(
  # define boundaries to calculate - it can also be only one boundary
  boundary = c("bluewater", "greenwater", "lsc", "nitrogen", "biosphere"),
  config_scenario = config_path_scenario,
  config_reference = config_path_reference,
  spatial_scale = "global",
  time_span_scenario = time_span_scenario,
  time_span_reference = time_span_reference,
  # moving average over nyear_window years
  time_series_avg = nyear_window,
  approach = list("bluewater" = "porkka2024",
                  "greenwater" = "porkka2024",
                  "nitrogen" = "schulte_uebbing2022"),
  # boundary specific parameters, see individual boundary functions for details
  time_span_baseline = time_span_scenario, # for biosphere integrity
  path_baseline = paste0(path_outputs, "/pnv/"), # for biosphere integrity
  savanna_proxy = list(vegc = 7500) # for forest biome definition in the land system change boundary
)

### plot timeseries

# There are three options for plotting the global status timeseries:
# a) one timeseries panel for each boundary control variable, in a Cartesian
#    coordinate system
# b) all timeseries plotted in one panel, in a Cartesian coordinate system
# c) one timeseries panel for each boundary control variable, in a polar
#    coordinate system (called "stylized")

# a) one timeseries panel for each boundary control variable
plot_status(
  x = global_status,
  stylized = FALSE,
  filename = "./global_timeseries_panels.png"
)

# b) all timeseries plotted in one panel
plot_status(
  x = global_status,
  stylized = FALSE,
  all_in_one = TRUE,
  filename = "./global_timeseries_all_in_one.png"
)

# c) one timeseries panel for each boundary control variable, in a polar
#    coordinate system
plot_status(
  x = global_status,
  stylized = TRUE,
  filename = "./global_timeseries_stylized.png"
)

```

#### **Example** *Compare two calculation approaches for the same boundary*  

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# In this example, only the bluewater boundary is calculated, for two different
# approaches. The results are then jointly plotted in a Cartesian coordinate
# system.

# define analysis and reference period
time_span_scenario <- as.character(1901:2017)
time_span_reference <- as.character(1500:1699)

# Aprroach A: Following Porkka et al. (2024) (https://doi.org/10.1038/s44221-024-00208-7)
# Referring to global area with discharge deviations outside the pre-industrial
# range

bluewater_status_porkka <- calc_status(
  # define boundaries to calculate - it can also be only one boundary
  boundary = c("bluewater"),
  config_scenario = config_path_scenario,
  config_reference = config_path_reference,
  spatial_scale = "global",
  time_span_scenario = time_span_scenario,
  time_span_reference = time_span_reference,
  # no moving average, by setting the number of years to average over to 1:
  time_series_avg = 1,
  approach = list("bluewater" = "porkka2024")
)

# Aprroach B: Following Rockström et al. (2009) (https://doi.org/10.1038/461472a)
# Referring to global bluewater consumption, but adapting the defaullt boundary
# and high risk values

bluewater_status_rockstroem <- calc_status(
  # define boundaries to calculate - it can also be only one boundary
  boundary = c("bluewater"),
  config_scenario = config_path_scenario,
  config_reference = config_path_reference,
  spatial_scale = "global",
  time_span_scenario = time_span_scenario,
  time_span_reference = time_span_reference,
  # no moving average, by setting the number of years to average over to 1:
  time_series_avg = 1,
  approach = list("bluewater" = "rockstroem2009"),
  # change default planetary boundary and high risk values, following Gerten
  # et al. 2013 (https://doi.org/10.1016/j.cosust.2013.11.001),
  # all referring to km3/yr as defined in `metric_files.yml`
  thresholds = list("bluewater" = list(holocene = 0,
                                       pb = 2800,
                                       high_risk = 4000))
)

# plot both results in one timeseries panel
# merge both results into one list
# TODO test if this is working!
bluewater_status <- list("bluewater" = bluewater_status_porkka$bluewater,
                         "bluewater" = bluewater_status_rockstroem$bluewater)

plot_status(
  x = bluewater_status,
  stylized = FALSE,
  all_in_one = TRUE,
  filename = "./global_bluewater_status_porkka_vs_rockstroem.png"
)

```

### Gridded status\

#### **Example** *Gridded status calculation and plotting*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}


# define analysis and reference period
time_span_scenario <- as.character(2008:2017)
time_span_reference <- as.character(1500:1699)
# if the land use effect is to be isolated, the time span for the reference
# period can be set to the same as the scenario period:
# time_span_reference <- as.character(2008:2017)
# This way, climate change induced changes are excluded.

# calc status at the gridded level
gridded_status <- calc_status(
  boundary = c("lsc", "nitrogen", "bluewater", "biosphere"),
  config_scenario = config_path_scenario,
  config_reference = config_path_reference,
  time_span_scenario = time_span_scenario,
  time_span_reference = time_span_reference,
  spatial_scale = "grid",
  # set time_series_avg to NULL, to not calculate a timeseries, but to average
  # over the entire scenario time span
  time_series_avg = NULL,
  # boundary specific parameters, see individual boundary functions for details
  time_span_baseline = time_span_scenario, # for biosphere integrity
  path_baseline = paste0(path_outputs, "/pnv/"), # for biosphere integrity
  savanna_proxy = list(vegc = 7500) # for forest biome definition in the land system change boundary
)

# plot status at the gridded level, for details see `?status_maps`
# There are two options for plotting a map with the gridded status(es):
# a) plot the control variable status of each boundary (e.g. deforestation share
#   for the land system change boundary)
# b) plot the risk level of each boundary, based on a normalized color scale
#    and the boundary and high risk values

# a) control variable status
plot_status(
  x = gridded_status,
  filename = "./gridded_status_control_variable.png",
  grid_path = paste0(path_outputs, "/pnv/grid.bin.json"),
  risk_level = FALSE
)

# b) risk level
plot_status(
  x = gridded_status,
  filename = "./gridded_status_risk_level.png",
  grid_path = paste0(path_outputs, "/pnv/grid.bin.json"),
  risk_level = TRUE
)

```

&nbsp;
&nbsp;
&nbsp;


\newpage
### Regional status\
For the regional status, the calculation is performed on a aggregated subglobal
level that makes sense for the respective boundary. For example, for the land
system change boundary, the regional status is calculated at the level of forest
biomes, for biosphere integrity at the level of biomes, and for bluewater at
the level of river basins.

####  **Example** *Regional status calculation and plotting*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}

# define analysis and reference period
time_span_scenario <- as.character(2008:2017)
time_span_reference <- as.character(1500:1699)

# calc status at the regional level
regional_status <- calc_status(
  boundary = c("lsc", "bluewater", "greenwater", "biosphere"),
  config_scenario = config_path_scenario,
  config_reference = config_path_scenario,
  time_span_scenario = time_span_scenario,
  time_span_reference = time_span_reference,
  spatial_scale = "subglobal",
  approach = list("bluewater" = "porkka2024",
                  "greenwater" = "porkka2024"),
  path_baseline = paste0(path_outputs, "/pnv/"), # for biosphere integrity
  time_span_baseline = time_span_scenario, # for biosphere integrity
  savanna_proxy = list(vegc = 7500), # for forest biome definition in the land system change boundary
)

# plot status at the regional level

# As for the gridded status, there are two options for plotting a map with the
# regional status(es): directly the control variable status or the risk level.

# risk level status
plot_status(
  x = regional_status,
  filename = "./regional_status_risk_level.png",
  grid_path = paste0(path_outputs, "/pnv/grid.bin.json"),
  risk_level = TRUE
)

```

&nbsp;
&nbsp;

\newpage

### Notes & tips

- Some boundary status calculations may take long (particularly for green and
  bluewater based on the approach following Porkka et al. 2024).
  To speed up the calculation, the `in_parallel` argument can be set to `TRUE`
  in the `calc_status` function to parallelize some of the calculations.
- It is advisable to submit the calculations as a job to slurm

